# Kin-finding lab

In this session, we are going to get our hands dirty with a lovely data set
of genetic data from 1,484 bearded seals and use those data to find kin pairs
from amongst the ${1,484 \choose 2} = 1,100,386$ possible pairs that can be formed
from all those samples.  

This data set includes 2,569 genetic markers.  These markers were filtered down
for quality and reliability from a much larger set.  We won't talk about that
process, here, but keep in mind that a huge part of a successful CKMR project
involves getting and curating a quality set of genetic markers.

We will first estimate how much power we expect to have for identifying
kin pairs with this marker set, and then we will actually do the kin-finding.
Both of these steps will be done using the R package 'CKMRsim' that Eric wrote.
It is available from GitHub at https://github.com/eriqande/CKMRsim and its
online documentation can be read at: https://eriqande.github.io/CKMRsim/.

If you want to use 'CKMRsim' yourself in a different project, you would need
to install it like this:
```{r, eval=FALSE}
remotes::install_github("eriqande/CKMRsim")
```
However, we have already included it in the `tws-ckmr-2022` project using
'renv', so you don't have to install it now.

Before proceeding, we will load all of the libraries that we will be using
for this session.  Additionally, we will download and install the
binary of the `mendel` program so that we can simulate genetic markers with
physical linkage.
```{r inst-pkgs, message=FALSE}
library(tidyverse)
library(CKMRsim)

# the next step tests to see if mendel was already installed.
# if not, it installs it
if(system.file("bin", package = "CKMRsim") == "") {
  install_mendel(Dir = system.file(package = "CKMRsim"))
}
```

## Looking at the genetic data and pivoting it 

We load the genetic data from an RDS file:
```{r load-genos}
beardo_genos <- read_rds("data/beardo-genos.rds")
```

Here are the first 10 rows 9 columns of that data set:
```{r cols11}
beardo_genos[1:10, 1:9]
```
The full data set has `r nrow(beardo_genos)` rows and `r ncol(beardo_genos)` columns.

This is a standard genotype format. The first column gives the sample name and
every subsequent pair of columns gives the alleles found in the sample at a
genetic marker or "locus".

In order to manipulate these more easily and prepare them for CKMRsim we will
pivot them into a longer format.
```{r pvl}
long_genos <- beardo_genos %>%
  rename(Indiv = Our_sample) %>%
  pivot_longer(
    cols = -Indiv, 
    names_to = c("Locus", "gene_copy"), 
    names_sep = "\\.", 
    values_to = "Allele"
  )
```

This looks like:
```{r pvlview}
long_genos
```


## Kin finding power analysis assuming no physical linkage

To assess the power of a set of markers for kin finding, CKMRsim requires
the allele freqencies in a particular format.  These steps
calculate the allele frequencies and format them as required:
```{r}
afreqs_ready <- long_genos %>%
  count(Locus, Allele) %>%
  group_by(Locus) %>%
  mutate(Freq = n / sum(n),
         Chrom = "Unk",
         Pos = as.integer(factor(Locus, levels = locus_info$Locus))) %>%
  ungroup() %>%
  select(Chrom, Pos, Locus, Allele, Freq) %>%
  arrange(Pos, desc(Freq)) %>%
  mutate(AlleIdx = NA,
         LocIdx = NA) %>%
  filter(!is.na(Allele)) %>%
  reindex_markers()
```

The allele freqs look like this:
```{r}
afreqs_ready
```

These allele frequencies then get compiled into an R object
that includes the results of a number of calculations that account for genotyping
error and a number of matrices that allow rapid simulation of kin pairs from the
data.

This takes a little while because it is a fairly large
data set.  But not more than a minute on my laptop.
```{r create_ckmr, cache=TRUE}
ckmr <- create_ckmr(
  D = afreqs_ready,
  kappa_matrix = kappas[c("PO", "FS", "HS", "HAN", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
```

The `kappa_matrix` specifies the identify coefficients (the $\kappa$'s) for
the different relationships we want.  The remaining options tell it to
use a fairly generic model for genotyping error with an error rate (`epsilon`) at
which we expect about 1% of the genotypes to have a genotyping error.

The variable `kappas` is a piece
of package data that looks like this:
```{r kappaview}
kappas
```
So, the line 
```r
kappas[c("PO", "FS", "HS", "HAN", "U"), ]
```
is merely picking out the rows from that matrix that we want to focus on: we
want to be prepared to do calculations concerning the relationships of:

- `U`: unrelated
- `HAN`: half aunt-niece (which includes half uncle-niece and half uncle-nephew, etc.)
- `HS`: half sibling
- `FS`: full sibling
- `PO`: parent-offspring


Now we can simulate some log likelihood ratios and plot them to see how they look.
```{r simQs}
# This simulated a large number of pairwise genotype probabilites
# at each locus
Qs <- simulate_Qij(
  ckmr, 
  calc_relats = c("PO", "FS", "HS", "HAN", "U"),
  sim_relats = c("PO", "FS", "HS", "HAN", "U") 
)


# this lets us extract those genotype probabilities in different
# ways to calculate a variety of different log-likelihood ratios
# for different relationships.  

# in this particular case, we are looking at log likelihood ratios
# for the hypothesis of Parent-Offspring vs Unrelated
PO_U_logls <- extract_logls(
  Qs,
  numer = c(PO = 1),
  denom = c(U = 1)
)

# And we can plot the distribution of those logl ratios for each
# of the different true relationships.
ggplot(
  PO_U_logls,
  aes(x = logl_ratio, fill = true_relat)
) +
  geom_density(alpha = 0.25)
```

Clearly, PO and U should be easily resolved.  And it turns out that physical linkage does 
not affect the distribution of log-likelihood ratios when the truth relationship is
either U or PO.  And we can use importance sampling to estimate false positive rates.
```{r}
mc_sample_simple(
  Qs,
  nu = "PO"
)
```

Yep! No chance, whatsoever, of an unrelated pair being mistaken as a PO pair.

For all other relationship types, the fact of physical linkage is important.
In particular, physical linkage really becomes important when we start to consider
kin pairs that are less related than half-siblings, but not a lot less.  The
super relevant category here is half-aunt neice.  

## Power for kin-finding while accounting for physical linkage

With CKMRsim, we can simulate likelihood ratios (which are calculated without
reference to the physical linkage) under the influence of physical linkage,
**BUT** In order to do that formally, we need to know where in the genome all these
markers are.  We don't know that, in this case, but we can still get a very good sense for the effect of
linkage by imagining that all these markers are spread throughout an appropriately sized
genome.  Since we are simulating lots of pairs in this way, the exact details of the physical
linkage are not quite as important as if we were trying to calculate likelihoods given the
linkage.

### Make a pseudo genome for these critters

A quick search for the bearded seal shows they have a Karyotype of 2N=34. 
That means they have, 16 pairs of autosomes and one pair of sex chromosomes.
That is not a huge abundance of chromosomes, but it is not as bad as fruit flies.
If you look at pictures of the karyotype, the smallest chromosomes are about 1/4
the size of the largest.

There is a draft genome out there that has it at about 2.4 gigabases.  Cool.  We
can use that to create a pseudo-genome and then sprinkle our markers into it
in order to get a sense for how much physical linkage will affect the distribution
of the log likelihood ratios.

```{r, fig.height=1}
# make fake chromosome lengths
fake_chromo_lengths <- geometric_chromo_lengths(15, 2.4, 0.25)

# here is a plot of what they look like
fake_chromo_lengths$chrom_length_plot
```

Now that we have that approximate genome to play with, let's go ahead and randomly
place our variants into it.
```{r}
set.seed(5)
afreqs_link <- sprinkle_markers_into_genome(afreqs_ready, fake_chromo_lengths$chrom_lengths)
```


Now we can make a new CKMR object that has this (fake) physical location data for
the markers.
```{r}
ckmr_link <- create_ckmr(
  D = afreqs_link,
  kappa_matrix = kappas[c("PO", "FS", "HS", "HAN", "U"), ],
  ge_mod_assumed = ge_model_TGIE,
  ge_mod_true = ge_model_TGIE,
  ge_mod_assumed_pars_list = list(epsilon = 0.005),
  ge_mod_true_pars_list = list(epsilon = 0.005)
)
```

And, now, to simulate with physical linkage, we need to include the pedigree information.


```{r}
Qs_link_BIG <- simulate_Qij(
  ckmr_link, 
  calc_relats = c("PO", "FS", "HS", "HAN", "U"),
  sim_relats = c("PO", "FS", "HS", "HAN", "U"),
  unlinked = FALSE, 
  pedigree_list = pedigrees
)
```

In fact. let's look at the PO/U and the FS/U logl ratios here:
```{r}
Qs_link_BIG %>%
  extract_logls(numer = c(PO = 1), denom = c(U = 1)) %>%
  ggplot(aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25) +
  ggtitle("PO/U Logl Ratio")
```


Aha!  When we simulate those likelihood ratios whilst taking account of
physical linkage, we see that there is a lot more spread in the distributions,
and, consequently, a whole lot more overlap.

Now, let us look at things when we use other likelihood ratios:

```{r}
Qs_link_BIG %>%
  extract_logls(numer = c(FS = 1), denom = c(U = 1)) %>%
  ggplot(aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25) +
  ggtitle("FS/U Logl Ratio")
```



```{r}
Qs_link_BIG %>%
  extract_logls(numer = c(PO = 1), denom = c(FS = 1)) %>%
  ggplot(aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25) +
  ggtitle("PO/FS Logl Ratio")
```

```{r}
Qs_link_BIG %>%
  extract_logls(numer = c(FS = 1), denom = c(HS = 1)) %>%
  ggplot(aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25) +
  ggtitle("FS/HS Logl Ratio")
```

```{r}
Qs_link_BIG %>%
  extract_logls(numer = c(HS = 1), denom = c(HAN = 1)) %>%
  ggplot(aes(x = logl_ratio, fill = true_relat)) +
  geom_density(alpha = 0.25) +
  ggtitle("HS/HAN Logl Ratio")
```

For this last one, let's see how many Unrelated might conceivably
get in there:
```{r}
mc_sample_simple(
  Q = Qs,
  Q_for_fnrs = Qs_link_BIG,
  nu = "HS",
  de = "HAN",
  method = "IS"
)
```

Yep.  Clearly no unrelateds will get in there.


Looking at all of these is quite informative.  It suggests that the order that
we will want to do this in is:

1. Find all PO pairs by high PO/U and also PO/FS > 0.
2. Once we have those, we will look at FS/HS and look at the
distribution there for things above -20 or so.  Obviously no
unrelateds will be anywhere near that, but we should still check with
the importance sampling.
3. Once we have the PO and FS individuals we will compute the HS/HAN
logl_ratios and then set the cutoff at what we calculated before, with
an FNR of about 20%.  But we will want to investigate the empirical
distribution of all those values, too, to see if we can see a HAN bump.



## Doing the pairwise comparisons

Because of the ages of these individuals it is probably likely that many of them could
not be PO are FS.  Many probably would be more likely to be GP than HS, all things
being equal, too.  Some of these might be AN instead of HS.  

But, regardless, we can tackle these things in a few different steps.

### Make sure we don't have duplicate sample in here

Unless there is a lot of wonky inbreeding going on, it is unlikely that samples
that are not _the same individual_ will have close to 100% matching genotypes.

Here we find all the pairs that mismatch at no 500 or fewer loci out of the 2500
that are in the data set:
```{r}
matchers <- find_close_matching_genotypes(
  LG = long_genos,
  CK = ckmr,
  max_mismatch = 500
)
matchers
```

There are none.  Cool.


### I need a section here showing the syntax of the pairwise_kin_logl_ratios() function

Boing!

### Doing pairwise comparisons somewhat systematically

So, we only really have four columns to make: PO/U, PO/FS, FS/HS, and HS/HAN.
I am going to make them all and then put them into a big data frame. The
number of markers is the same every time.
```{r}
pw_4_LRTs <- lapply(
  X = list(
    POU = c("PO", "U"),
    POFS = c("PO", "FS"),
    FSHS = c("FS", "HS"),
    HSHAN = c("HS", "HAN")
  ),
  FUN = function(x) {
    pairwise_kin_logl_ratios(
      D1 = long_genos, 
      D2 = long_genos, 
      CK = ckmr,
      numer = x[1],
      denom = x[2],
      num_cores = 8
    )
  }
) %>%
  bind_rows(
  .id = "lr_type"
) %>%
  pivot_wider(names_from = lr_type, values_from = logl_ratio)


```

And now we can see how things compare to our simulations:
```{r}
topPO <- pw_4_LRTs %>%
  arrange(desc(POU)) %>%
  filter(POU > 0)

topPO
```
And from that we see the first 5 look like solid PO pairs, and the 8th row
has some wonky thing going on. I thought at first that maybe it was inbred
full siblings, but then it would have a higher FSHS.  Weird.  I guess I
will call all 6 of those PO, but seeing this gives me some doubts about
how consistent the genotype data are.
```{r}
likelyPO <- topPO %>%
  filter(POFS > 20)
```

And then we go on to the FSHS and we want to first look at the distribution
for values > -20:
```{r}
topFS <- pw_4_LRTs %>%
  anti_join(likelyPO, by = c("D2_indiv", "D1_indiv")) %>%
  arrange(desc(FSHS)) %>%
  filter(FSHS > -20)
topFS
```

So, that is two individuals that look like solid full siblings  and two that
are not super solid, but may be more likely than being HS, unless there
are a boatload of HSs.  So, for now, we will call these the likely FS. But, again,
there are intimations that the genetic data may not be super consistent.

So, finally let's start looking for half-siblings.
```{r}
remaining <- pw_4_LRTs %>%
  anti_join(bind_rows(likelyPO, topFS), by = c("D2_indiv", "D1_indiv"))
```

We can have a look at the bumps we might see on the extreme right end of the
unrelated distribution and then the others.
```{r}
gHSHAN <- ggplot(remaining, aes(x = HSHAN)) + 
  geom_histogram()
```

Plot all of them:
```{r}
gHSHAN +
  ggtitle("All Remaining Pairs")
```

Now, have a look at the extreme right edge there:
```{r}
gHSHAN +
  xlim(-55, NA) +
  ggtitle("Pairs with HSHAN > -55")
```

That makes sense, but we still can't see the other bumps.  Let's
go for everything above -25:

```{r}
gHSHAN +
  xlim(-25, NA) +
  ggtitle("Pairs with HSHAN > -25")
```

OK, let's also look at this another way:
```{r}
set.seed(5)
ggplot(
  remaining %>% filter(HSHAN > -55) %>% mutate(y = 1),
  aes(x = HSHAN, y = y)
) +
  geom_jitter(width = 0, height = 0.2, colour = "blue") +
  ylim(0,2)

```

That is pretty telling.  We have all the unrelateds, then a cluster of
likely HAN, and then an impressive break before the likely HS, and then
one pair that is clearly not a half-sib pair. Let's look at estimated
FNRs and FPR here:
```{r}
mc_sample_simple(
  Qs_link_BIG,
  nu = "HS",
  de = "HAN",
  tr = "HAN",
  method = "vanilla",
  lambda_stars = c(-2:20))
```
So the false positive rate here when the truth is HAN and HSHAN > 9 is only .0109: about
1 out of 100.  And there aren't anywhere near that many HAN's (just not that many blue dots).
So, if we set our cutoff at 9, we can be pretty confident that there will be no
HANs in there.  

On the other side, the highest one in the lower (HAN) cluster is about -2.  That would
give us a FNR of .03.  Since there are only about 19 HS's we probably don't have any
false negatives.  But we can still use a FNR or .03 or .04.

The one dot that is out at 87 or something is likely a wonky full sib.

So we can form our final assignments:
```{r}
hs_part <- remaining %>%
  filter(HSHAN > 0) %>%
  mutate(assigned = ifelse(HSHAN < 50, "HS", "FS*"))  # it gets a star cuz it is super wonky

final_pairs <- bind_rows(
  likelyPO %>% mutate(assigned = ifelse(POU>385, "PO", "PO*")),
  topFS %>% mutate(assigned = ifelse(FSHS > 10, "FS", "FS*")),
  hs_part
) %>%
  arrange(POU, POFS, FSHS, HSHAN)

final_pairs
```

And now we should go and check this with the meta data to see how these
square with that.  Keeping in mind that the HS categories could be
grandparent-grandchild, or, much less likely, full-aunt-niece.

The 18 half-siblings seem pretty solid to me.


